Metacircular Virtualization \& \\ Practical Nock Interpretation

### 2025-08-13 mss.tex
Add tombstone.
Add refs

In 1958, John McCarthy began his lifelong work on the Lisp programming language \citep{McCarthy1996}.  While the form would evolve over the years, the fundamental components included S-expressions, or parenthesized lists of homoiconic code–data; \texttt{'} or quote, a deferred expression; and \texttt{eval}, a function to evaluate quoted code.\footnote{Paul Graham, in his 2001 short essay ``What Made Lisp Different'', grouped these concepts into the phrase ``The whole language always available'' \citep{Graham2001}.}  Together these components allowed McCarthy and colleague Steve Russell to implement a Lisp interpreter in Lisp itself, the first metacircular interpreter \citep{McCarthy1978}.  Such was a radical departure from the imperative programming paradigm of the time, exemplified in Fortran (1954–55); Lisp's reliance on recursion and higher-order functions was both a precursor to the functional program paradigm and a rejuvenation of the Church and Curry schools of fundamental computing logic.
%McCarthy1996 https://www-formal.stanford.edu/jmc/history/lisp/node2.html
%Graham2001 https://paulgraham.com/diff.html
%McCarthy1978 McCarthy, John; Wexelblat, Richard L. (1978). History of programming languages. Association for Computing Machinery. pp. 173–183.

@@

Amin and Rompf (\citeyear{Amin2017}) analyzed how metacircular interpreters build on top of each other, each adding a new layer of abstraction.  While primarily examining a Lisp dialect, they showed practically how interpreted language extensions (even in cascading towers) can be collapsed to bottom-level interpretation, similar to Nock's own \lstinline[style=inlinecode]{++mink} function.  They concisely epitomized a line of research on reflective languages and self-interpreted languages, such as quasiquotation in Lisp \citep{Bawden1999} and Template Meta-Haskell \citep{Sheard2002}.  The ultimate point of this enquiry was to show that metacircularity was a powerful and well-grounded tool for building interpreters, and that it could be used to in fact build practical interpreters.
%Amin2017 https://dl.acm.org/doi/10.1145/3158140
%Bawden1999 https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=3ad9e27f0a4b1847655ba41acf1b40f8f401366b
%Sheard2002 Sheard, Tim; Jones, Simon Peyton (3 October 2002). "Template meta-programming for Haskell". Proceedings of the 2002 ACM SIGPLAN workshop on Haskell. Association for Computing Machinery. pp. 1–16. doi:10.1145/581690.581691. ISBN 1581136056. S2CID 6096655.  https://dl.acm.org/doi/10.1145/581690.581691

@@

\begin{itemize}
  \item  \lstinline[style=inlinecode]{++mink}, compute bottom-level virtual Nock.
  \item  \lstinline[style=inlinecode]{++mock}, compute formula on subject with hint.  Wraps \lstinline[style=inlinecode]{++mink}; common entrypoint.
  \item  \lstinline[style=inlinecode]{++mule}, kick a trap (deferred computation).  Wraps \lstinline[style=inlinecode]{++mock}; common entrypoint.
  \item  \lstinline[style=inlinecode]{++soft}, wrap a typecast as a unit (i.e., fail with \lstinline[style=inlinecode]{~} rather than crash).
\end{itemize}

@@

The remainder of this section analyzes how each \lstinline[style=inlinecode]{++mink} implements each Nock opcode, including particularly the fake opcode 12.  The following code samples are taken from \lstinline[style=inlinecode]{/sys/hoon} in the Urbit codebase.

@@

Opcode 1 is the constant function, which returns a constant value extricated from the formula.  This is implemented in \lstinline[style=inlinecode]{++mink} as follows:

@@

Hearkening back to \citet{Amin2017}'s collapsed towers of interpreters, the affordances of \lstinline[style=inlinecode]{++mock} as a metacircular interpreter permit us to build a tower of interpreters on top of it without incurring excessive performance overhead.  As the author of \texttt{u3.md} concluded, ``we [even] simply treat Nock proper as a special case of \lstinline[style=inlinecode]{mock}.''  The conventional \lstinline[style=inlinecode]{++mink} gate is built in Hoon, but what would happen if we stripped it down and built it in Nock itself with the minimal necessary machinery?  What would such an interpreter look like, and how would it relate to actual operation on a runtime in practice?

Let us extend \lstinline[style=inlinecode]{++mink} to implement a logical loobean operator.  (Call the modified gate \lstinline[style=inlinecode]{++pink}.)  This operation will be invoked with a new fake opcode 13.  Under the hood, opcode 13 will simply supply logically negate a value (via an opcode 6), much as how opcode 12 supplies a consistent Nock formula for the scry operation.  (Similar to \lstinline[style=inlinecode]{[0 0]}, we will be able to use \lstinline[style=inlinecode]{[13 1 2]} to initiate a crash if desired.)

\begin{lstlisting}[style=listingcode]
:: *[a 13 b]      NOT *[a b]
  [%13 subject=*]
=/  argument  $(formula argument.formula)
?.  ?=(%0 -.argument)  argument
?^  product.argument  [%2 trace]
?:  =(%1 product.argument)  [%0 %.y]
?:  =(%0 product.argument)  [%0 %.n]
[%2 trace]
\end{lstlisting}

@@

Metacircular virtualization is a powerful tool for building interpreters, and Nock is not only no exception, it exemplifies the practice.  The design of \lstinline[style=inlinecode]{++mock} and its underlying \lstinline[style=inlinecode]{++mink} function provides a practical way to interpret Nock in a way that is both efficient and extensible.  One could imagine, for instance, a Nock interpreter cousin to \lstinline[style=inlinecode]{++mock} that is instrumented for partial evaluation of formulas (as in a debugger or syntax highlighter robust to syntax errors).  Nock supports a rich capacity to build metacircular interpreters as a first-class feature, a potential which remains as yet relatively unexplored in the Nock ecosystem.

@@

@@

%   :-  (jink [42 [0 1]])
%   (jink [42 [0 0]])

@@

@@

>>>>

In 1958, John McCarthy began his lifelong work on the Lisp programming language \citep{McCarthy1996}.  While the form would evolve over the years, the fundamental components included S-expressions, or parenthesized lists of homoiconic code–data; \texttt{'} or quote, a deferred expression; and \texttt{eval}, a function to evaluate quoted code.\footnote{Paul Graham, in his 2001 short essay ``What Made Lisp Different'', grouped these concepts into the phrase ``The whole language always available'' \citep{Graham2001}.}  Together these components allowed McCarthy and colleague Steve Russell to implement a Lisp interpreter in Lisp itself, the first metacircular interpreter \citep{McCarthy1978}.  Such was a radical departure from the imperative programming paradigm of the time, exemplified in Fortran (1954–55); Lisp's reliance on recursion and higher-order functions was both a precursor to the functional program paradigm and a rejuvenation of the Church and Curry schools of fundamental computing logic.

@@

Amin and Rompf (\citeyear{Amin2017}) analyzed how metacircular interpreters build on top of each other, each adding a new layer of abstraction.  While primarily examining a Lisp dialect, they showed practically how interpreted language extensions (even in cascading towers) can be collapsed to bottom-level interpretation, similar to Nock's own \lstinline[style=inlinecode]{++mink} function.  They concisely epitomized a line of research on reflective languages and self-interpreted languages, such as quasiquotation in Lisp \citep{Bawden1999} and Template Meta-Haskell \citep{Sheard2002}.  The ultimate point of this enquiry was to show that metacircularity was a powerful and well-grounded tool for building interpreters, and that it could be used to in fact build practical interpreters.

@@

\begin{itemize}
  \item  \lstinline[style=inlinecode]{++mink}, compute bottom-level virtual Nock.
  \item  \lstinline[style=inlinecode]{++mock}, compute formula on subject with hint.  Wraps \lstinline[style=inlinecode]{++mink}; common entrypoint.
  \item  \sloppy \lstinline[style=inlinecode]{++mule}, kick a trap (deferred computation).  Wraps \lstinline[style=inlinecode]{++mock}; common entrypoint.
  \item  \lstinline[style=inlinecode]{++soft}, wrap a typecast as a unit (i.e., fail with \lstinline[style=inlinecode]{~} rather than crash).
\end{itemize}

@@

\sloppy
The remainder of this section analyzes how each \lstinline[style=inlinecode]{++mink} implements each Nock opcode, including particularly the fake opcode 12.  The following code samples are taken from \lstinline[style=inlinecode]{/sys/hoon} in the Urbit codebase.

@@

\sloppy
Opcode 1 is the constant function, which returns a constant value extricated from the formula.  This is implemented in \lstinline[style=inlinecode]{++mink} as follows:

@@

Hearkening back to \citeauthor{Amin2017}'s collapsed towers of interpreters, the affordances of \lstinline[style=inlinecode]{++mock} as a metacircular interpreter permit us to build a tower of interpreters on top of it without incurring excessive performance overhead.  As the author of \texttt{u3.md} concluded for the Vere C runtime, ``we [even] simply treat Nock proper as a special case of \lstinline[style=inlinecode]{mock}.''  The conventional \lstinline[style=inlinecode]{++mink} gate is built in Hoon, but what would happen if we stripped it down and built it in Nock itself with the minimal necessary machinery?  What would such an interpreter look like, and how would it relate to actual operation on a runtime in practice?

Let us extend \lstinline[style=inlinecode]{++mink} to implement a logical loobean operator.  (Call the modified gate \lstinline[style=inlinecode]{++pink}.)  This operation will be invoked with a new fake opcode 13.  Under the hood, opcode 13 will simply logically negate a value (via an opcode 6), much as how opcode 12 supplies a consistent Nock formula for the scry operation.  (Similar to \lstinline[style=inlinecode]{[0 0]}, we will be able to use \lstinline[style=inlinecode]{[13 1 2]} to initiate a crash if desired.)

\begin{lstlisting}[style=listingcode]
:: *[a 13 b]                    NOT *[a b]
  [%13 subject=*]
=/  argument  $(formula argument.formula)
?.  ?=(%0 -.argument)  [%2 trace]
?^  product.argument  [%2 trace]
?:  =(%1 product.argument)  [%0 %.y]
?:  =(%0 product.argument)  [%0 %.n]
[%2 trace]
\end{lstlisting}

@@

Metacircular virtualization is a powerful tool for building interpreters, and Nock is not only no exception, it exemplifies the practice.  The design of \lstinline[style=inlinecode]{++mock} and its underlying \lstinline[style=inlinecode]{++mink} function provides a practical way to interpret Nock in a way that is both efficient and extensible.  One could imagine, for instance, a Nock interpreter cousin to \lstinline[style=inlinecode]{++mock} that is instrumented for partial evaluation of formulas (as in a debugger or syntax highlighter robust to syntax errors).  Nock supports a rich capacity to build metacircular interpreters as a first-class feature, a potential which remains as yet relatively unexplored in the Nock ecosystem. \tombstone

@@

% \begin{lstlisting}[style=listingcode]

@@

%   :: :-  (jink [42 [0 1]])
%   (jink [42 [13 0 1]])

@@

%     :: *[a 13 b]
%       [%13 argument=*]
%     ?.  ?=(%0 -.argument)  (dec argument)
%     ?^  product.argument  [%2 trace]
%     ?:  =(%1 product.argument)  [%0 %.y]
%     ?:  =(%0 product.argument)  [%0 %.n]
%     [%2 trace]

@@

% \end{lstlisting}
